Syntax and Semantics of Cedille

### 2021-12-09 docs/semantics/paper.tex
docs: semantics: paper: update author emails
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>
docs: semantic: first sketch of refinement types proposal
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>

\author{Aaron Stump, Chris Jenkins \\
  Computer Science \\
  The University of Iowa \\
\texttt{aaron-stump@uiowa.edu}}

@@

\begin{figure}
  \centering
  \[
    \begin{array}{lll}
      \infer{\vdash \cdot}{}
      & \infer{
        \vdash \Gamma,x:T
        }{\Gamma \vdash T \tpsynth \star}
      & \infer{
        \vdash \Gamma,X:\kappa
        }{
        \Gamma \vdash \kappa
        }
    \end{array}
  \]
  \caption{Rules for checking a context is well-formed}
  \label{fig:ctxt}
\end{figure}

@@

>>>>

\author{Aaron Stump, Chris Jenkins \\
  Computer Science \\
  The University of Iowa \\
  \texttt{aaron-stump@uiowa.edu},
  \texttt{christopher-jenkins@uiowa.edu}}

@@

\begin{figure}
  \centering
  \[
    \begin{array}{lll}
      \infer{\vdash \cdot}{}
      & \infer{
        \vdash \Gamma,x:T
        }{\vdash \Gamma
          \quad \Gamma \vdash T \tpsynth \star}
      & \infer{
        \vdash \Gamma,X:\kappa
        }{
        \vdash \Gamma
        \quad \Gamma \vdash \kappa
        }
    \end{array}
  \]
  \caption{Rules for checking a context is well-formed}
  \label{fig:ctxt}
\end{figure}

@@

\section{Refinement types proposal}

The following is a proposal to replace dependent intersections with refinement
types.

\begin{figure}[h!]
  \centering
  \[
    \begin{array}{c}
      \begin{array}{cc}
        \infer{
        \Gamma \vdash \{ x : T_1\ |\ T_2\} \tpsynth \star
        }{
        \Gamma \vdash T_1 \tpsynth \star
        \quad \Gamma,x : T_1 \vdash T_2 \tpsynth \star
        }
        &
          \infer{
           \{ x : T_1\ |\ T_2\} \cong^{\text{t}} \{ x : S_1\ |\ S_2\}
          }{
           T_1 \cong S_1 \quad T_2 \cong S_2
          }
      \\ \\
        \infer{
         \Gamma \vdash [ t_1\ |\ t_2] \tpcheck \{ x : T_1\ |\ T_2\}
        }{
         \Gamma \vdash t_1 \tpcheck T_1
         \quad \Gamma \vdash t_2 \tpcheck [t_1/x]T_2
        }
        &
          \infer{
           \Gamma \vdash \uparrow t \tpsynth T_1
          }{
           \Gamma \vdash t \tpsynth \{ x : T_1\ |\ T_2\}
          }
      \end{array}
      \\ \\
      \infer{
       \Gamma \vdash \epsilon\ t_1\ -\ y.t_2 \Leftrightarrow T
      }{
      \begin{array}{c}
        \Gamma \vdash t_1 \tpsynth \{ x : T_1\ |\ T_2\}
        \quad y \notin \mathit{FV}(|t_2|)
        \\
        y \notin \mathit{FV}(T)
        \quad \Gamma, y : T_2d[\uparrow t_1/x] \vdash t_2 \Leftrightarrow T
      \end{array}
      }
      \\ \\
      \begin{array}{lllllll}
        |[ t_1\ |\ t_2]| & = & |t_1| &\
        & |\uparrow t| & = & |t|
        \\ |\epsilon\ t_1\ -\ y.t_2| & = & |t_2|
      \end{array}
      \\ \\
      \begin{array}{lll}
        \interp{\{ x : T_1\ |\ T_2 \}}_{\sigma,\rho}
        & =
        & \{E_1 \in \interp{T_1}_{\sigma,\rho}\ |\ \exists E_2 \in
          \interp{T_2}_{\sigma[x \mapsto \choice(E)],\rho}\}
      \end{array}
    \end{array}
  \]
  \caption{Refinement types}
  \label{fig:refinement-types}
\end{figure}

\begin{proof}[Theorem~\ref{thm:snd}]
  \startcase{.2cm}
  \[
    \infer{
      \Gamma \vdash \{ x : T_1\ |\ T_2\} \tpsynth \star
    }{
      \Gamma \vdash T_1 \tpsynth \star
      \quad \Gamma, x : T_1 \vdash T_2 \tpsynth \star
    }
  \]

  The set $\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}$ is explicitly defined to
be a subset of $\interp{T_1}_{\sigma,\rho}$, which is in $\mathcal{R}$,
by the IH applied to the first premise.  Since for any
$A\subseteq\mathcal{L}$, $[A]_{\cbe}$ is in $\mathcal{R}$, to show that
$\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}$ is also in $\mathcal{R}$ it suffices
to show definedness of $\interp{T_2}_{\sigma[x\mapsto \choice(E)],\rho}\}$
(which is used in the predicate picking out the
particular subset of $\interp{T_1}_{\sigma,\rho}$), for
$E\in\interp{T_1}_{\sigma,\rho}$.  For such $E$, $\choice(E)$ is defined
(since $\interp{T_1}_{\sigma,\rho}\in\mathcal{R}$ and hence $E\in\interp{T_1}_{\sigma,\rho}$ is nonempty)
and in $E$,
so $\sigma[x\mapsto\choice(E)]\in\interp{\Gamma,x:T_1}$.
So by the IH for the second premise,
$\interp{T_2}_{\sigma[x\mapsto\choice(E),\rho]}$ is defined.

\startcase{.2cm}
\[
  \infer{\Gamma\vdash [ t_1\ |\ t_2 ] \tpcheck \{ x : T_1\ |\ T_2\}}
  {
    \Gamma \vdash t_1 \tpcheck T_1
    \quad \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
  }
\]
By the IH, we have $[\sigma |t_1|]_{\cbe}\in\interp{T_1}_{\sigma,\rho}$ and
$[\sigma |t_2|]_{\cbe}\in\interp{[t_1/x]T_2}_{\sigma,\rho}$.
By Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
$[\sigma |t_1|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice([\sigma
  |t_1|]_{\cbe}),\rho}$.
Thus, we exhibit \([\sigma |t|]_{\cbe}\) for the existentially quantified
\(E_2\) in the definition of \(\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}\).
So, we see that \([\sigma|t_1|] = [\sigma|[ t_1\ |\ t_2]|] \in \interp{\{x :
  T_1\ |\ T_2\}}_{\sigma,\rho}\), as desired.

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \uparrow t \tpsynth T}{\Gamma\vdash t \tpsynth \{ x : T_1\
    |\ T_2\}}
\]
The desired conclusion follows easily from the IH and the semantics of
refinement types.

\startcase{.2cm}
\[
  \infer{
    \Gamma \vdash \epsilon\ t_1\ -\ y.t_2 \Leftrightarrow T
  }{
    \begin{array}{c}
      \Gamma \vdash t_1 \tpsynth \{ x : T_1\ |\ T_2\}
      \quad y \notin \mathit{FV}(|t_2|)
      \\
      y \notin \mathit{FV}(T)
      \quad \Gamma, y : [\uparrow t_1/x]T_2 \vdash t_2 \Leftrightarrow T
    \end{array}
  }
\]

By the IH on the first premise, \([\sigma|t_1|] \in \interp{\{x : T_1\ |\
  T_2\}}\).
This means that there exists \(E_2 \in \interp{T_2}_{\sigma[x \mapsto
  \choice([\sigma|t_1|])]}\).
By Lemma~\ref{lem:termsubstinterp}, \(E_2 \in \interp{[\uparrow
  t_1/x]T_2}_{\sigma,\rho}\).
Thus, \((\sigma \uplus [y \mapsto \choice(E_2),\rho]) \in \interp{\Gamma,y :
  [\uparrow t_1/x]T_2}\), and by the IH on the fourth premise \(\sigma[y \mapsto
\choice(E_2)]|t_2| \in \interp{T}_{\sigma[y \mapsto \choice(E_2)],\rho}\).
Since \(y \notin \mathit{FV}(|t_2|)\) and \(y \notin \mathit{FV}(T)\), we have
\(\sigma|t_2| = \sigma|\epsilon\ t_1\ -\ y.t_2| \in \interp{T}_{\sigma,\rho}\)
as desired.

\subsection{Dependent intersections}

\begin{figure}[h]
  \centering
  \[
    \begin{array}{lll}
      \abs{\iota}{x}{T_1}{T_2}
      & =
      & \{ x : T_1\ |\ \{y : T_2\ |\ \{y \simeq x \}\} \}
      \\ {[ t_1 , t_2 ]}
      & =
      & {[ t_1\ |\ [t_2\ |\ \beta]]}
      % & =
      % & [ t_1\ |\ [ t_2\ |\ \beta ]]
      \\ t.1
      & =
      & \uparrow t
      \\ t.2
      & =
      & \epsilon\ t\ -\ y. \epsilon\ y\ - z. \varphi\ z\ -\ (\uparrow y) \{t\}
    \end{array}
  \]
  \caption{Dependent intersections}
  \label{fig:refinement-iota}
\end{figure}

Figure~\ref{fig:refinement-iota} shows an encoding of dependent intersections
using refinement types.
Abusing notation, let \(\mathcal{E}\) be the natural extension of these rules to
types, kinds, contexts, and terms.

\begin{theorem}
  \label{thm:iota-as-refinement}
  \ \\
  \begin{itemize}

  \item Let \(u\) be an untyped \(\lambda\)-calculus term.
    If \(|t| = u\) then \(|\mathcal{E}(t)| = u\)

  \item \(\mathcal{E}([t/x]T) = [\mathcal{E}(t)/x]\mathcal{E}(T)\)
  \item 
    \begin{enumerate}
    \item If \(\vdash \Gamma\) then \(\vdash \mathcal{E}(\Gamma)\)
    \item If \(\Gamma \vdash \kappa\) then \(\mathcal{E}(\Gamma) \vdash \mathcal{E}(\kappa)\)
    \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\mathcal{E}(\Gamma) \vdash
      \mathcal{E}(T) \tpsynth \mathcal{E}(\kappa)\)
    \item If \(\Gamma \vdash t \Leftrightarrow T\) then \(\mathcal{E}(\Gamma) \vdash
      \mathcal{E}(t) \Leftrightarrow \mathcal{E}(T)\)
    \end{enumerate}
  \end{itemize}
\end{theorem}

\begin{proof}
  We show a few interesting cases.

  \startcase{.2cm}
  \[
    \mathit{If}\ \Gamma \vdash [ t_1 , t_2 ] \tpcheck \abs{\iota}{x}{T_1}{T_2}\ 
    \mathit{then}\ \mathcal{E}(\Gamma) \vdash \mathcal{E}([ t_1 , t_2]) \tpcheck
    \mathcal{E}(\abs{\iota}{x}{T_1}{T_2})
  \]

  We have \(|t_1| =_{\beta\eta} |t_2|\) from the fourth premise of the assumed
  derivation.
  By the erasure lemma, this means \(|\mathcal{E}(t_1)| =_{\beta\eta}
  |\mathcal{E}(t_2)|\).
  Appeal to the IH on the two typing premises of the derivation to obtain
  \(\mathcal{D}_1 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_1) \tpcheck
  \mathcal{E}(T_1)\) and
  \(\mathcal{D}_2 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_2) \tpcheck
  \mathcal{E}([t_1/x]T_2)\).
  By the substitution lemma, this means \(\mathcal{E}(t_2)\) checks against type
  \([\mathcal{E}(t_1)/x]\mathcal{E}(T_2)\).
  We build the following derivation.
  \[
    \infer{
      \mathcal{E}(\Gamma) \vdash [ \mathcal{E}(t_1)\ |\ [\mathcal{E}(t_2)\ |\
      \beta ]] \tpcheck \{ x : \mathcal{E}(T_1)\ |\ \{y : \mathcal{E}(T_2)\ |\
      \{y \simeq x\}\}\}
    }{
      \mathcal{D}_1 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_1) \tpcheck
      \mathcal{E}(T_1)
      \quad
      \infer{
        \mathcal{E}(\Gamma) \vdash [ t_2\ |\ \beta ] \tpcheck
        \{ y : T_2\ |\ \{y \simeq x\}\}
      }{
        \mathcal{D}_2 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(T_2) \tpcheck
        [\mathcal{E}(t_1)/x]\mathcal{E}(T_2)
        \quad
        \infer{
          \mathcal{E}(\Gamma) \vdash \beta \tpcheck \{\mathcal{E}(t_1) \simeq \mathcal{E}(t_2)\}
        }{
          |\mathcal{E}(t_1)| = |\mathcal{E}(t_2)|
        }
      }
    }
  \]

  \startcase{.2cm}
  \[
    \mathit{If}\ \Gamma \vdash t.1 \tpsynth T\ \mathit{then}\ \mathcal{E}(\Gamma) \vdash
    \mathcal{E}(t.1) \tpsynth \mathcal{E}(T)
  \]

  Appeal to the IH on the premise of
  the assumed typing derivation to obtain \(\mathcal{D} :: \mathcal{E}(\Gamma) \vdash
  \mathcal{E}(t) \tpsynth \mathcal{E}(\abs{\iota}{x}{T_1}{T_2})\).
  By the definition of \(\mathcal{E}\) we therefore have that \(\mathcal{E}(t)\)
  synthesizes type \(\{ x : \mathcal{E}(T_1)\ |\ \{ y : \mathcal{E}(T_2)\ |\ \{
  y \simeq x\}\}\}\).
  The goal is to derive \(\mathcal{E}(\Gamma) \vdash \mathcal{E}(t.1) \tpsynth
  \mathcal{E}(T_1)\).
  \[
    \infer{
      \mathcal{E}(\Gamma) \vdash\ \uparrow \mathcal{E}(t) \tpsynth \mathcal{E}(T_1)
    }{
      \mathcal{D} :: \mathcal{E}(\Gamma) \vdash t \tpsynth
      \{ x : \mathcal{E}(T_1)\ |\ \{ y : \mathcal{E}(T_2)\ |\ \{ y \simeq x \}\}\}
    }
  \]
\end{proof}
